Nitric oxide feedback to ciliary photoreceptor cells orchestrates UV avoidance

### 2023-05-26 Manuscript/Manuscript.Rmd
MS: edits to discussion, NO in working memory

Nitric oxide (NO) is one such modulator with unique physical and signalling properties.
This free radical synthesized by nitric oxide synthase (NOS) is short-lived and can diffuse across biological membranes [@Cudeiro1999; @thomas2015].
Canonical NO signalling involves the activation of neuronal NOS following calcium increase, NO production and diffusion, and the NO-dependent activation of soluble guanylate cyclases (sGC) in target cells leading to cGMP production [@Bredt1990; @Hoelscher1997].
Given that NOS activation is calcium dependent and NOS shows neuron-type-specific expression [@Gibbs1998a; @Wildemann1999; @Mobley2022], NO action can lead to the activity-dependent modulation of neural circuits at specific sites [@Wang2007; @Vielma2014; @jacoby2018] .

@@

We aimed next to identify the NO-receptor in the cPRCs.
NO generally acts via soluble guanylate cyclases (sGC), belonging to the guanylate cyclase family with a CYC domain (PFAM domain: PF00211).
NO binding to the heme group of sGC leads to increased cyclic guanosine monophosphate (cGMP) production.
Analysis of sGCs in *Platynereis* indicated that these genes are not expressed in any of the cells of the cPRC circuit [@Veraszto2017].
Recently, Moroz and coworkers reported an atypical but widely conserved family of guanylyl cyclases with a NIT (nitrite/nitrate sensing) domain (PF08376) (NIT-GC) as potential mediators of NO signalling [@Moroz2020].
To identify NIT-GCs in *Platynereis*, we searched transcriptome resources and retrieved 15 potential NIT-GC homologs (Figure 4---figure supplement 1 and 2).
To analyse the relationship of these sequences to metazoan NIT-GCs, we retrieved protein sequences with a CYC domain from the transcriptome and genome databases of 45 metazoan and 2 choanoflagellate species.
We carried out cluster analysis and did phylogenetic reconstruction on a group of membrane-bound guanylyl cyclases with sGCs as an outgroup.
In agreement with Moroz et al. [@Moroz2020], we found a group of GCs with NIT domains with representatives in placozoans, cnidarians, some ecdysozoans, echinoderms, and lophotrochozoans.
The 15 *Platynereis* sequences belonged to several deeply diverged clades in the phylogenetic tree (Figure 4---figure supplement 1 and 2).
To characterise the expression of NIT-GCs, we used previously published spatially mapped single-cell transcriptome data [@Achim2015; @williams2017].
Among the 15 NIT-GCs, two showed high and specific expression in the cPRCs and one was expressed in the INNOS cells (Figure 4---figure supplement 2).
In the single-cell data, we could identify the cPRCs by the specific expression of *c-opsin1* and the *pedal-peptide2 neuropeptide precursor* (*MLD proneuropeptide*), previously described cPRC markers [@Arendt2004; @williams2017] (Figure 4---figure supplement 3A).
The INNOS cells were identified by NOS expression and spatial mapping in the brain [@Achim2015].
We decided to focus on two NIT-GCs expressed in the cPRCs and with a full-length sequence, *NIT-GC1* and *NIT-GC2*.
To confirm the single-cell data, we first carried out in situ hybridisation chain reaction (HCR) with probes for *NIT-GC1* and *NIT-GC2* mRNA. Both genes were specifically expression in the four cPRCs, as confirmed by co-labeling with an acetylated α-tubulin antibody and with an HCR probe against *pedal peptide 2/MLD proneuropeptide* (Figure 4C, D and Figure 4---figure supplement 3A-C).
To analyse the subcellular localisation of NIT-GC1 and NIT-GC2 at the protein level, we raised and affinity-purified polyclonal antibodies against a specific peptide sequence from both proteins.
In immunostainings, we found that NIT-GC1 was localise to the region corresponding to the axonal projections of the cPRCs in the anterior nerve plexus (Figure 4E).
Co-immunostaining with the rabbit NIT-GC1 antibody and a custom rat antibody raised against *Platynereis* NOS revealed the localisation of both proteins in close proximity in the neurosecretory plexus (Figure 4---figure supplement 3D).
In contrast, NIT-GC2 specifically labelled the ramified sensory cilia of the cPRCs (Figure 4F).
These different subcellular localisations suggest that the two NIT-GCs are involved in different intracellular signalling processes in the ciliary and axonal regions of the cPRCs.

@@

NO is a free radical with a millisecond-to-second scale half life in vivo and a limited signalling range.
In neuronal signalling, NOS and sGC have been found to localise at close proximity at synapses [@burette2002][@garthwaite2015a].
We identified 12 sGCs in *Platynereis*, but none of these is expressed in the cPRC.
Instead, we identified an unconventional NIT-domain containing GC, NIT-GC1 as the mediator of NO retrograde signalling.
The NIT domain was first identified in bacteria and animal NIT-GCs have only been recently reported [@Shu2003; @Moroz2020].
Bacterial NIT domains regulate cellular functions in response to changes in extracellular and intracellular nitrate and/or nitrite concentrations [@camargo2007].
NO is readily converted to nitrate and nitrite [@santos2011; @moeller2019; @garthwaite2015a] and these molecules accumulate in placozoans and cnidarians in cells and tissues with high NOS activity [@Moroz2004; @Moroz2020].
It is possible that the NIT domains in NIT-GCs may also sense nitrate and nitrite, as in bacteria.
Furthermore, if they show different sensitivities to NO, nitrite and nitrate, it is possible that differences in their half-lives could give rise to a range of activation timings [@lundberg2011].
We have found 15 NIT-GCs in *Platynereis*, suggesting a wide range of functions.
In addition, NIT-GC1 and NIT-GC2 showed very different subcellular localisation and functions.
It is not clear whether the functional differences are due to the different localisation or also to differences in biochemical function.
Overall, the diversification and spatially different localisation patterns of NIT-GCs may increase their signalling repertoire.
Our data revealed key roles for two *Platynereis* NIT-GCs in phototraonsduction and NO-mediated neuromodulation suggesting that these molecules contribute to the diversity of neuronal signalling.
Sponges and ctenophores, where NO signalling is present, lack NIT-GCs.
In placozoans as many as 12 have been found, and cnidarians also contain a large diversity.

**Nitric oxide feedback modulation of UV-avoidance behaviour**

Why does a seemingly roundabout feedback pathway exist?
The activity of synaptic circuits is richly modulated monoamines, neuropeptides and gaseous molecules.
The gaseous second messenger nitric oxide (NO) is known to activate a post- to presynaptic retrograde signalling cascade involving cyclic guanosine monophosphate (cGMP) production.
It has been reported that NO produced in one of the mammalian retinal cells modulates cGMP-mediated neurotransmission by activating the presynaptic nerve sGC (a general target of NO) through retrograde signalling [@Vielma2014; @Wei2012].
Retrograde signalling of NO is also known to be one of the molecules that regulate memory formation.
In Drosophila, NOS of PPL1-γ1pedc shortens memory retention while promoting fast memory updating in response to new experiences, indicating that NO regulates forgetting [@Aso2019].
Because NO has a short half-life and may be a suitable second messenger for encoding working memories whose duration must be limited.
R-neuron inputs to the ellipsoid in Drosophila also represent visual object orientation for tens of seconds after the disappearance of an object, suggesting that they are involved in short-term visual memory dependent on nitric oxide signalling [@Kuntz2017].
In this study, our mathematical modelling results focused on the time of UV stimulation and showed that the late activity comes more delayed during short UV and sooner during long UV.
This suggests that animals may have a UV length-dependent short-term memory system that regulates cPRC activation, and that there may be mechanisms by which UV stimulation is enhanced and maintained such that NO feedback maintains neural activation even with slight UV.
It is possible that these mechanisms may have been early signalling pathways controlling stimulus changes across space and time in animals' short-term memory.
In planarian UV avoidance behaviour, different time and length scales of neuropeptide and small molecule transmission were reported to generate incoherent patterns of neural activity that competitively control behaviour and memory [@bray2023].
Short-term memory is essential for locating less exposed areas to minimise UV-induced damage in animals.

@@

This work was funded by the Wellcome Trust (214337/Z/18/Z). This project has received funding from the European Research Council (ERC) under the European Union’s Horizon 2020 research and innovation programme (grant agreement No 101020792). KJ has been supported by a JSPS Postdoctoral fellowhip, LAYG by a BBSRC Discovery fellowship (BB/W010305/1).

>>>>

Nitric oxide (NO) is one such modulator with unique physical and signalling properties.
This free radical synthesized by nitric oxide synthase (NOS) from L-arginine is short-lived and can diffuse across biological membranes [@Cudeiro1999; @thomas2015].
Canonical NO signalling involves the Ca2+/calmodulin-dependent activation of NOS following calcium increase, NO production and diffusion, and the NO-dependent activation of soluble guanylate cyclases (sGC) cell-autonomously or in other cells leading to cGMP production [@Bredt1990; @Hoelscher1997].
Given that NOS activation is calcium dependent and NOS shows neuron-type-specific expression [@Gibbs1998a; @Wildemann1999; @Aso2019; @Mobley2022], NO action can lead to the activity-dependent modulation of neural circuits at specific sites [@Wang2007; @Vielma2014; @jacoby2018; @Aso2019] .

@@

We aimed next to identify the NO-receptor in the cPRCs.
NO generally acts via soluble guanylate cyclases (sGC), belonging to the guanylate cyclase family with a CYC domain (PFAM domain: PF00211).
NO binding to the heme group of sGC leads to increased cyclic guanosine monophosphate (cGMP) production.
Analysis of sGCs in *Platynereis* indicated that these genes are not expressed in any of the cells of the cPRC circuit [@Veraszto2017].
Recently, Moroz and coworkers reported an atypical but widely conserved family of guanylyl cyclases with a NIT (nitrite/nitrate sensing) domain (PF08376) (NIT-GC) as potential mediators of NO signalling [@Moroz2020].
To identify NIT-GCs in *Platynereis*, we searched transcriptome resources and retrieved 15 potential NIT-GC homologs (Figure 4---figure supplement 1 and 2).
To analyse the relationship of these sequences to metazoan NIT-GCs, we retrieved protein sequences with a CYC domain from the transcriptome and genome databases of 45 metazoan and 2 choanoflagellate species.
We carried out cluster analysis and did phylogenetic reconstruction on a group of membrane-bound guanylyl cyclases with sGCs as an outgroup.
In agreement with Moroz et al. [@Moroz2020], we found a group of GCs with NIT domains with representatives in placozoans, cnidarians, some ecdysozoans, echinoderms, and lophotrochozoans.
The 15 *Platynereis* sequences belonged to several deeply diverged clades in the phylogenetic tree (Figure 4---figure supplement 1 and 2).
To characterise the expression of NIT-GCs, we used previously published spatially mapped single-cell transcriptome data [@Achim2015; @williams2017].
Among the 15 NIT-GCs, two showed high and specific expression in the cPRCs and one was expressed in the INNOS cells (Figure 4---figure supplement 2).
In the single-cell data, we could identify the cPRCs by the specific expression of *c-opsin1* and the *pedal-peptide2 neuropeptide precursor* (*MLD proneuropeptide*), previously described cPRC markers [@Arendt2004; @williams2017] (Figure 4---figure supplement 3A).
The INNOS cells were identified by NOS expression and spatial mapping in the brain [@Achim2015].
We decided to focus on two NIT-GCs expressed in the cPRCs and with a full-length sequence, *NIT-GC1* and *NIT-GC2*.
To confirm the single-cell data, we first carried out in situ hybridisation chain reaction (HCR) with probes for *NIT-GC1* and *NIT-GC2* mRNA. Both genes were specifically expression in the four cPRCs, as confirmed by co-labeling with an acetylated α-tubulin antibody and with an HCR probe against *pedal peptide 2/MLD proneuropeptide* (Figure 4C, D and Figure 4---figure supplement 3A-C).
To analyse the subcellular localisation of NIT-GC1 and NIT-GC2 at the protein level, we raised and affinity-purified polyclonal antibodies against a specific peptide sequence from both proteins.
In immunostainings, we found that NIT-GC1 was localise to the region corresponding to the axonal projections of the cPRCs in the anterior nerve plexus (Figure 4E).
Co-immunostaining with the rabbit NIT-GC1 antibody and a custom rat antibody raised against *Platynereis* NOS revealed the localisation of both proteins in dots in the neurosecretory neuropil (Figure 4---figure supplement 3D).
In contrast, NIT-GC2 specifically labelled the ramified sensory cilia of the cPRCs (Figure 4F).
These different subcellular localisations suggest that the two NIT-GCs are involved in different intracellular signalling processes in the ciliary and axonal regions of the cPRCs.

@@

NO is a free radical with a millisecond-to-second scale half life in vivo and a limited signalling range.
In neuronal signalling, NOS and sGC have been found to localise at close proximity at synapses [@burette2002][@garthwaite2015a].
We identified 12 sGCs in *Platynereis*, but none of these is expressed in the cPRC.
Instead, we identified an unconventional NIT-domain containing GC, NIT-GC1 as the mediator of NO retrograde signalling.
The NIT domain was first identified in bacteria and animal NIT-GCs have only been recently reported [@Shu2003; @Moroz2020].
Bacterial NIT domains regulate cellular functions in response to changes in extracellular and intracellular nitrate and/or nitrite concentrations [@camargo2007].
NO is readily converted to nitrate and nitrite [@santos2011; @moeller2019; @garthwaite2015a] and these molecules accumulate in placozoans and cnidarians in cells and tissues with high NOS activity [@Moroz2004; @Moroz2020].
NIT domains in NIT-GCs may also sense nitrate and nitrite, as in bacteria.
Furthermore, if different NIT-GCs show different sensitivities to NO, nitrite and nitrate, it is possible that differences in the half-lives of these molecules could give rise to a range of activation timings [@lundberg2011].
We have found 15 NIT-GCs in *Platynereis*, suggesting a wide range of functions.
In addition, NIT-GC1 and NIT-GC2 showed very different subcellular localisation and functions.
It is not clear whether the functional differences are due to the different localisation or also to differences in biochemical function.
Overall, the diversification and spatially different localisation patterns of NIT-GCs may increase their signalling repertoire.
Our data revealed key roles for two *Platynereis* NIT-GCs in phototransduction and NO-mediated neuromodulation suggesting that these molecules contribute to the diversity of neuronal signalling.
Sponges and ctenophores, where NO signalling is present, lack NIT-GCs.
In placozoans as many as 12 have been found, and cnidarians also contain a large diversity.
Soluble GCs are expressed broadly in the *Drosophila* and mammalian brain and the specificity of NO signalling seems to depend on localised NOS expression and activation, and ectopic NOS expression can confer NO-dependent effects [@Aso2019].
In *Platyneries* larvae

**Nitric oxide confers short-term memory to circuit activity**

Why does a seemingly roundabout feedback pathway exist?
The activity of synaptic circuits is richly modulated monoamines, neuropeptides and gaseous molecules.
The gaseous second messenger nitric oxide (NO) is known to activate a post- to presynaptic retrograde signalling cascade involving cyclic guanosine monophosphate (cGMP) production.
It has been reported that NO produced in one of the mammalian retinal cells modulates cGMP-mediated neurotransmission by activating the presynaptic nerve sGC (a general target of NO) through retrograde signalling [@Vielma2014; @Wei2012].
Retrograde signalling of NO is also known to be one of the molecules that regulate memory formation.
In Drosophila, NOS of PPL1-γ1pedc shortens memory retention while promoting fast memory updating in response to new experiences, indicating that NO regulates forgetting [@Aso2019].
In the ellipsoid body of the *Drosophila* central brain, NOS is specifically expressed in the R3 ring neuron, but not other ring (R) neurons [@Kuntz2017]. R3 neurons are required for visual working memory with NO signalling mediating a short-term (~4 sec) component of the visual memory of objects. In this circuit, NO produced in R3 axons acts in an autocrine manner to activate sGC in R3 axons leading to a CNG-dependent (cyclic nucleotide-gated ion channel) temporary increase in calcium levels [@Kuntz2017]. Because of the short life-time of NO, this molecule may be well suited to encode transient memory traces [@Kuntz2017].

In this study, our mathematical modelling results focused on the time of UV stimulation and showed that the late activity comes more delayed during short UV and sooner during long UV.
This suggests that animals may have a UV length-dependent short-term memory system that regulates cPRC activation, and that there may be mechanisms by which UV stimulation is enhanced and maintained such that NO feedback maintains neural activation even with slight UV.
It is possible that these mechanisms may have been early signalling pathways controlling stimulus changes across space and time in animals' short-term memory.
In planarian UV avoidance behaviour, different time and length scales of neuropeptide and small molecule transmission were reported to generate incoherent patterns of neural activity that competitively control behaviour and memory [@bray2023].
Short-term memory is essential for locating less exposed areas to minimise UV-induced damage in animals.

@@

This work was funded by the Wellcome Trust (214337/Z/18/Z).
This project has received funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme (grant agreement No 101020792).
KJ has been supported by a JSPS Postdoctoral fellowhip, LAYG by a BBSRC Discovery fellowship (BB/W010305/1).
